---
title: "Assignment 3: Data Exploration"
author: "Chenjia Liu"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Check your working directory, load necessary packages (tidyverse, lubridate), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r Load package and data}
getwd()
library(tidyverse)
library(lubridate)

Neonics<- read.csv("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv",stringsAsFactors = TRUE)
Litter<- read.csv("./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv",stringsAsFactors = TRUE)

```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer:Neonicotinoids are active substances used in plant protection products to control harmful insects. Neonicotinoids transfuse into pollen grains, nectar, cell sap and the food synthesized in the plants treated by them. These chemicals are highly neurotoxic in their mode of action to insects and other arthropods and act on the nicotinic acetylcholine receptor

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer:Litter and Woody debris are important parts of forest and stream ecosystems because it has a role in carbon budgets and nutrient cycling, is a source of energy for aquatic ecosystems, provides habitat for terrestrial and aquatic organisms, and contributes to structure and roughness, thereby influencing water flows and sediment transport

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1.Liter and fine woody debris sampling is executed at terrestrial NEON sites that contain woody vegetation >2m
tall
 2.One litter trap pair (one elevated trap and one ground trap) is deployed,either radomly or targated denpends on the aerial percent coverage,  for every 400 m2 plot area, resulting in 1-4 trap pairs per plot.
 3. Ground traps are sampled once per year. Target sampling frequency for elevated traps varies by vegetation present at the site.

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r Check dimention of Neonics}
dim(Neonics)
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?

```{r investigate the most common effects}
colnames(Neonics)
summary(Neonics$Effect)
```

> Answer:The most common effects are population and mortality. We are specifically intertested in these effects because we want to know how the presense of Neonics affect the population (reduce population), and the size of population is related to mortality, which could increase as the result of Neonics. 


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: The `sort()` command can sort the output of the summary command...]

```{r Determine the most commonly studied species }
summary(Neonics$Species.Common.Name)
sort(summary(Neonics$Species.Common.Name))
```

> Answer:The 6 most commonly studied species in the dataset from most to least are: 1. Honey Bee 2.  Parasitic Wasp 3. Buff Tailed Bumblebee 4. Carniolan Honey Bee 5. Bumble Bee 6. Italian Honeybee. We are interested in bees becuase they are the major pollinator for the agriculture, and they are greatly affected by the use of systemic pesticides such as Neonics. Our food security and resources is strongly related to the presence of bees. 

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric?

```{r Check the class of Conc.1..Author.}
class(Neonics$Conc.1..Author.)
view(Neonics$Conc.1..Author.)
```

> Answer:It is a factor. It is not numeric because there are not only number in this column, but there are other objects, such as 'NR'

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r Plot the number of studies conducted by publication year }
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year), bins = 50)
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r Plot the number of studies conducted by publication year and different Test.Location}
ggplot(Neonics) +
  geom_freqpoly(aes(x = Publication.Year,  color = Test.Location), bins = 50)
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer:The most common location before 2000 is field natural. But after 2000, the lab became the most common location and the count of publication has also increase a lot. The publication reached peak between 2010 and 202. 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r bar graph of Endpont counts}
ggplot(Neonics) +
  geom_bar(aes(x = Endpoint))+  
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

> Answer:The most common endpoints are LOEL and NOEL. LOEL:Lowest-observable-effect-level: lowest dose (concentration) 
producing effects that were significantly different (as reported 
by authors) from responses of controls (LOEAL/LOEC)
NOEL:No-observable-effect-level: highest dose (concentration) 
producing effects not significantly different from responses of 
controls according to author's reported statistical test 
(NOEAL/NOEC)

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r  class of collectDate}
#class(Litter$collectDate) #It's a factor
Litter$collectDate <- as.Date(Litter$collectDate, format = "%Y-%m-%d")
class(Litter$collectDate)
view(Litter)
unique(Litter$collectDate)
```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r Count of plots sampled at Niwot Ridge}
unique(Litter$namedLocation)
summary(Litter$namedLocation)
```

> Answer:'Unique' is able to directly tell how many levels in the data, When using 'summary', we need to count number of levels but it also provide more information, ie. count of each location

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r bar graph of functionalGroup counts}
ggplot(Litter) +
  geom_bar(aes(x = functionalGroup))
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r boxplot and violin plot of dryMass by functionalGroup}
ggplot(Litter) +
  geom_boxplot(aes(x = functionalGroup, y = dryMass))
ggplot(Litter) +
  geom_violin(aes(x = functionalGroup, y = dryMass))
```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: In this case, only boxplot can show the IQR, median, and outliers, and the different length of the lines on both side of the box tell us the skewness of data. We can only see the range of our data from the violin plot in this case. 

What type(s) of litter tend to have the highest biomass at these sites?

> Answer:Needles tend to have the highest biomass at these sites. 
