---
title: "Assignment 2: Coding Basics"
author: "Chenjia Liu"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. 
Seq1_30 <- seq(1,30, 3) #Generate a sequence of numbers from one to 30, increasing by threes
#2. 
mean(Seq1_30)
median(Seq1_30) #Compute the mean and median of this sequence
#3. 
mean(Seq1_30)<median(Seq1_30)
```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
Student_names<-c("Tim","Tom", "John","Jack") #Character
class(Student_names)
Test_score<-c(27, 49,78, 91) #Numeric
class(Test_score)
Pass_or_not<-c(Test_score>50) #Logical
class(Pass_or_not)

Student_test_results<-data.frame("Student"=Student_names,"Score"=Test_score,"Passed"=Pass_or_not)
View(Student_test_results)
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer:The main difference is that matrices can only contain a single class of data, while data frames can consist of many different classes of data. In this case, the data frames has 3 different classes.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}
If_pass<-function(Test_score) {
  ifelse(Test_score>50,TRUE,FALSE)
}

If_pass2 <- function(Test_score) {
  if(Test_score < 50) {
    FALSE
  } 
  else {
    TRUE
  }
}
If_pass(Test_score)
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: `if` and `else` works well when your input is a value, but it won't work if the input is a vector. The 'ifelse" works in both cases. 
